Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/splitting.tex
Fix stray page break.
halfway through new 4.2

We now show how to translate terms $\colmix e A$ into the form $\pipeM c l r$,
pursuant to the correctness condition given in \ref{fig:splittingSummary}.
The algorithm is specified by the $\splitonesym$ relation (\cref{fig:termSplit}), 
which proceeds recursively on the structure of~$e$.

We start with splitting the unit value, $\tup{}$.
Masking tells us that units, being trivial, only split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
Plugging this in to the contract for splitting at \bbonem\ gives:
\[
	\text{if } \splitone {\tup {}} A c {l.r} \text{ then } \reduce c {\tup{\tup{},b}} \text{ and } [b/l]r \equiv \tup{}
\]
Although the only hard requirement for $b$, $c$, and $l.r$ is to satisfy these relations,
our actual choices are designed to do the least work possible.
In particular, we choose $b = \tup{}$; $c=\tup{\tup{},\tup{}}$; and $l.r = \_.\tup{}$.
That is, we choose a trivial boundary value.
All together, this yield the rule in \cref{fig:termSplit}.
This exercise proceeds much the same way for \texttt{grnd} blocks, 
with the purely first stage contents of the block standing in for one of the $\tup{}$s.

We next consider splitting a non-terminal, namely $\inl {e}$.
From the semantics, we already know that 
for some $\xi$ and $v$, $\diaone {e} \xi v$ and $\diaone {\inl e} \xi {\inl v}$.
Masking these outputs then gives $\rtab \xi v \vsplito \mval i q$ 
and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$.
Since splitting works from the inside out, we already know that
\[
	\splitone e A c {l.r} \text{ and } \reduce c {\tup{i,b}} \text{ and } [b/l]r \equiv q
\]
and what we need to do is find some $b'$, $c'$, and $l'.r'$ satisfying
\[
	\reduce {c'} {\tup{\inl i,b'}} \text{ and } [b'/l']r' \equiv q
\]
Once again, we take the simplest solution, which in this case is $b' = b$; $l'.r' = l.r$;
and $c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}$.  
This sort of technique generalize to all constructs which don't involve control flow.

\nr{working from here down}

>>>>

We now show how to translate terms $\colmix e A$ into the form $\pipeM c l r$,
pursuant to our splitting invariant.
The algorithm is specified by the $\splitonesym$ relation (\cref{fig:termSplit}), 
which proceeds recursively on the structure of~$e$. We derive a few cases here.

\paragraph {Unit} 
Consider splitting the unit value, $\tup{}$.
Masking says that units, being trivial, only split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
So in order for the split $\splitone {\tup {}} A c {l.r}$ to be valid, 
we need a $c$ and $l.r$ satisfying,
\[
	\reduce c {\tup{\tup{},b}} \text{ and } [b/l]r \equiv \tup{}
\]
Our algorithm chooses $c=\tup{\tup{},\tup{}}$ and $l.r = \_.\tup{}$.
Splitting $\pure{e}$ follows in similar fashion, 
where the immediate result is instead the result of evaluating $e$.

\paragraph {Injections} 
Now consider splitting a non-terminal, $\inl {e}$,
where $\diaone {e} \xi v$ and $\rtab \xi v \vsplito \mval i q$, 
and by the splitting invariant for $e$,
\[
	\splitone e A c {l.r} \text{ and } \reduce c {\tup{i,b}} \text{ and } [b/l]r \equiv q
\]
Since, $\diaone {\inl e} \xi {\inl v}$ and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$,
the validity of the split $\splitone {\inl e} A {c'} {l.r'}$ requires,
\[
	\reduce {c'} {\tup{\inl i,b'}} \text{ and } [b'/l]r' \equiv q
\]
Our algorithm chooses $c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}$ and $l.r' = l.r$.

\paragraph{Tuple} 
For an example with more than one argument, consider a tuple $\tup{e_1,e_2}$,
where $\diaone {e_1} \xi {v_1} \vsplito \mval {i_1} {q_1}$
and $\diaone {e_2} \xi {v_2} \vsplito \mval {i_2} {q_2}$.
The splitting invariants for $e_1$ and $e_2$ give
\[
	\splitone {e_i} A {c_i} {l_i.r_i} \text{ and } \reduce {c_i} {\tup{i_i,b_i}} \text{ and } [b_i/l_i]r_i \equiv q_i
\]
for $i=1,2$.

\paragraph {Case}
The most complicated rule is that of $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms, because of their divergent control flow.
We derive the correct splitting rule by considering the two possible execution paths seperately.
If the first branch was taken, then the predicate evaluated to $\diaone {e_1} {\xi_1} {\inl {v_1}}$ 
and the first branch evaluated to $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$.
Similarly these ouputs mask as $\rtab {\xi_1,\xi _2} {v_2} \vsplito \mval {i_2} {q_2}$ and $\rtab {\xi_1,\xi _2} {v_2} \vsplito \mval {i_2} {q_2}$.
Again because splitting works from the inside out, we already know that that the predicate obeys
\[
	\splitone {e_1} A {c_1} {l_1.r_1} \text{ and } \reduce {c_1} {\tup{\inl {i_1},b_1}} \text{ and } [b_1/l_1]r_1 \equiv q_1
\]
and the first branch obeys
\[
	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]c_2} {\tup{i_2,b_2}} \text{ and } [b_2/l_2]r_2 \equiv q_2
\]
